---
title: "Daily mail rebuttal"
output: github_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(cache = TRUE,
                      message = FALSE, 
                      warning = FALSE,
                      echo = FALSE)

library(pacman)
p_load(tidyverse, devtools, plotly, readxl)
```

## Accusations

The Daily Mail has made repeated accusations that PHE has made "blunders" in the reporting of case figures for the UK. In particular

-   The sum of daily reported cases is not equal to cumulative case count

-   Historical data is not corrected as a so-called expert thought it should be

-   There are lots of minor unexplained corrections

### Summary

The Daily Mail's accusations are unfounded and a sugnificant misintrepretation of the data, and PHE's role in it. They are based on:

-   An oversimplistic analysis

-   A failure to analyse the data correctly

-   A failure to understand key principles of public reporting

-   A failure to seek appropriate expert advice or review

## Reporting cases

To date more the 2.5 million people have tested positive for SARS-Cov-2 through PCR testing.

Cases are reported on people testing positive, **not** positive tests. Positive test results for England are held in a database known as SGSS. There are equivalent databases for Northern Ireland (NI), Scotland and Wales. These databases receive positive test results daily (usually overnight) from testing labs across the country.

The SGSS database collects data on all positive results for a wide range of communicable diseases, not only SARS-Cov-2. It has had to be substantially upgraded to manage lab returns for SARS-Cov-2 as the number of labs processing tests has grown. In the early part of the pandemic lab capacity was limited to NHS and PHE labs - known as *pillar1*. WIth the advent of mass-testing private and other labs were recruited - this tranche of testing is known as *pillar2*.

To calculate **people** tested the data needs to be de-duplicated, that is each person who has had a positive PCR test is counted only once no matter how many how many tests - positive or negative - they have had. This de-duplication exercise is conducted by Public Health England for England data, and respective public health agencies across the UK.

Although both pillar data were de-duplicated, they were de-duplicated separately because they had separate date streams - they weren't combined initially. In order to get an accurate count of the number of **people** who had tested positive pillar1 and pillar2 data needed to be combined and de-duplication conducted across the combined dataset. The results of this exercise resulted in major changes to the data published and was implemented in public reporting on 2nd July. Similar exercises were conducted by other nations in the days and weeks that followed.

Prior to 2nd July 2020 the public dashboard:

-   Reported a UK case figure including un-deduplicated pillar1 and pillar2 data for the 4 nations. This included the duplicated results for England - the largest contributor to the UK figure - with potentially multiple positive tests per person.

-   Reported England case data for pillar1 only. Case data for other countries was pillar1 with some pillar data for Scotland and Wales.

-   Was unable to report case numbers for sub-national areas like regions and local authorities.

The de-duplication exercise had a number of consequences:

1.  More than 30000 duplicate cases were removed from the reported UK total on 2nd July.

2.  More than 80000 pillar2 cases were added to the reported England total on 2nd July

3.  We were able to add complete case counts per local authority to the dashboard.

### Why doesn't the sum of daily reported cases add up the cumulative total reported cases?

There are 2 main reasons for this:

1.  We do **not correct the daily reported count. These figures are the reported historical record of the number of cases in the pandemic on the the day they are reported and are not retrospectively amended. Tens of thousands of people have kept paper or digital records of the daily reported figures since the beginning of the pandemic and would raise questions of trustworthiness if we retrospectively amended the data.**
2.  For this reason it there are corrections to be made, they are made to the cumulative total
3.  We **cannot** c**orrect the daily reported count.** If cases are removed or added to the database in batches we can know when the relevant tests were done but not when they would have been reported. We do not have the data to do this, and we only get aggregate data from the DAs. This was especially the case for the pillar2 data which reflected data over a four month period.
4.  This is unlike reporting by specimen data which is a better way to understand the pandemic (see below)
5.  For this reason, the UK cumulative figure is 30,000 less than the sum of daily reported cases since the 2nd July. A further correction was made on 27th November when 1,283 or so cases were removed from the UK total because of the correction made to reporting cases from the Milton Keynes Lighthouse lab.

Figure shows the differences between cumulative totals and the sum of the daily published totals for each nation and the UK as whole.

-   Panel A (UK) shows the loss of 30,000 + cases from the UK cumulative count plus the recent reduction from the Milton Keynes lab

-   Panel B (England) shows the addition of 80,000 cases to the England cumulative count plus the reduction of cases from the Milton Keynes lab

-   Panel C (NI) shows the gradual addition of 1000 cases

-   Panel D (Scotland) shows the addition of 2000 cases to the Scottish cumulative total

-   Panel E (Wales) shows the addition of the 800 or so pillar2 cases which weren't previously reported in the UK total but were added to the Welsh (and therefore the UK total) on 14th July. There have been multiple corrections to Welsh data since then.

### Specimen vs published data reporting

+-------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Comparison                                | Published date                                                                                                                                                                         | Specimen date                                                                                                                                                    |
+===========================================+========================================================================================================================================================================================+==================================================================================================================================================================+
| Daily figure                              | Yes (headline)                                                                                                                                                                         | No (data incomplete)                                                                                                                                             |
+-------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Available since the start of the pandemic | Yes                                                                                                                                                                                    | Yes but only since July 2nd                                                                                                                                      |
+-------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Reporting biases                          | Yes (under-reporting at weekends)                                                                                                                                                      | Less so                                                                                                                                                          |
+-------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Stability                                 | Daily figures are fixed                                                                                                                                                                | Dynamic - refreshed daily                                                                                                                                        |
+-------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Timeliness                                | Hard to say                                                                                                                                                                            | About 5-days out of data - more recent data is incomplete and unreliable                                                                                         |
+-------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Interpretability                          | Number of cases reported on a given day regardless of when tests were done or results obtained.                                                                                        | Number of cases reported for the test processed on any given day.                                                                                                |
+-------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Definition                                | The difference in the number of people tested positive in the database between 9 am on the day of publication and 9 am the day before regardless of when swabs were taken or processed | The number of cases detected from swabs processed on any given day. If someone has several positive swabs, the first positive test is taken as the specimen date |
+-------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------+

### The role of PHE

Much approbation is levelled at PHE. PHE does have a significant role in the reporting - it does collect English case data and produce death data, and members of PHE design, build and populate the public dashboard. on behalf of UK Government. Many others are involved however:

-   Approval and sign off for publishing is the DHSC and the SofS office.

-   Decisions on allocation or subtraction of cases is agreed with DHSC or SofS office

-   The DAs operate their own data collection and reporting rules and systems.

-   Metrics reported are determined by the govt., scientific advisers, CMOs office and PHE epidemiologists as well as SAGE, No 10 and DHSC and are influenced by WHO and international reporting.

```{r}

uk_data <- read_csv("https://api.coronavirus.data.gov.uk/v2/data?areaType=overview&metric=cumCasesByPublishDate&metric=cumCasesBySpecimenDate&metric=newCasesByPublishDate&metric=newCasesBySpecimenDate&format=csv")


nations_data <- read_csv("https://api.coronavirus.data.gov.uk/v2/data?areaType=nation&metric=cumCasesByPublishDate&metric=cumCasesBySpecimenDate&metric=newCasesByPublishDate&metric=newCasesBySpecimenDate&format=csv")



```

```{r}

cum_comparison <- function(df){
require(tidyverse)
return <- df %>%
 arrange(date) %>%
  mutate(date = lubridate::ymd(date)) %>%
  mutate_if(is.numeric, function(x) ifelse(is.na(x), 0, x)) %>%
  mutate(sumpub = sum(newCasesByPublishDate, na.rm = TRUE), 
         sumspec = sum(newCasesBySpecimenDate, na.rm = TRUE), 
         diffpub = sumpub - cumCasesByPublishDate,
         diffspec = sumspec - cumCasesBySpecimenDate, 
         cumpub = cumsum(newCasesByPublishDate), 
         cumspec = cumsum(newCasesBySpecimenDate)) %>%
  arrange(desc(date)) %>%
  mutate(diff_pub_1 = cumCasesByPublishDate - cumpub, 
         diff_spec_1 = cumspec - cumCasesBySpecimenDate) 



plot <- return %>%
  ggplot(aes(date, diff_pub_1)) +
  geom_col()

out <- list(return = return, plot = plot)
  
}



         
  
```

```{r}

uk <- cum_comparison(uk_data)

england_data <- nations_data %>%
  filter(str_detect(areaCode, "^E"))

england <- cum_comparison(england_data)

scotland_data <- nations_data %>%
  filter(str_detect(areaCode, "^S"))

scotland <- cum_comparison(scotland_data)

wales_data <- nations_data %>%
  filter(str_detect(areaCode, "^W"))

wales <- cum_comparison(wales_data)

ni_data <- nations_data %>%
  filter(str_detect(areaCode, "^N"))

ni <- cum_comparison(ni_data)

library(patchwork)

uk$plot + england$plot / ni$plot + scotland$plot / wales$plot +
  patchwork::plot_annotation(tag_levels = "A")



```

## 

### Reported vs specimen date

```{r}

change <- read_csv("~/Downloads/data_2021-Jan-04.csv")

change <- change %>%
  mutate(per = 100 * changeInNewCasesBySpecimenDate / 53180)
```

If we look at the 53,180 cases in England reported on 4th Janary 2020, `r round(change$per[1],2)`% of these were from tests performed on `r change$date[1]`; `r round(change$per[2],2)`% on `r change$date[2]`; `r round(change$per[3],2)`% on `r change$date[3]`; `r round(change$per[4],2)`% on `r change$date[4]` and `r round(change$per[5],2)`% on `r change$date[5]`. It takes at least 5 days for all the tests for any given day to be reported. Conversely, reported figures (cases by publication date) are the total of tests over many days and contrary to opinion do not give the most timely picture of the pandemic but an estimate of it.

Specimen date analysis is far more robust and although it is lagged by 5 days (data for 4th January is complete for 30th December), gives a much more robust picture of recent trends. The cumulative cases by specimen date is exactly equal to the sum of new cases by specimen date.
